---
title: "PR programming Assignment 3 week 4"
author: "Gustavo Ahumada"
date: "12/19/2020"
output: pdf_document
---
Programming Assignment 3
========================================================

Course 2 R Programming, Assignment 3 (Week 4), under Data Science by Johns Hopkins University


To begin, download data and unzip it into your R working directory.  

You can do  this in R with the following code:

```{r setup}
knitr::opts_knit$set(root.dir = '/Users/gusahu/Google Drive/Online_courses/r_programming/rprogramming_coursera/week_4') #set working directory
getwd()
```


```{r}
dataset <- getwd()
unzip("rprog_data_ProgAssignment3-data.zip", exdir = "hosp_compare")
```


# Part 1: Plot the 30-day mortality rates for heart attack

```{r, results = 'hide'}
outcome <- read.csv("hosp_compare/outcome-of-care-measures.csv", colClasses = "character", header = TRUE)
head(outcome, 10)
outcome[, 11] <- as.numeric(outcome[, 11], na.rm = TRUE) # set column 11
# NAs introduced by coercion
```

```{r}
hist(outcome[, 11], xlab = "Deaths", main = "Hospital 30-Day Death (Moratlity) Rates from Hart Attack", col = "lightblue")
```

# Part 2: Finding the best hospital in a state

```{r}
# create the function
best <- function(state, outcome) {
  ## Read outcome data
  data <- read.csv("hosp_compare/outcome-of-care-measures.csv", colClasses = "character", header = TRUE)
  data2 <- as.data.frame(cbind(data[, 2],   # hospital
                               data[, 7],   # state
                               data[, 11],  # heart attack
                               data[, 17],  # heart failure
                               data[, 23]), # pneumonia
                      stringsAsFactors = FALSE)
  colnames(data2) <- c("hospital", "state", "heart attack", "heart failure", "pneumonia")
  ## Check that state and outcome are valid
  if(!state %in% data2[, "state"]) {
    stop('invalid state')} else if(!outcome %in% c("heart attack", "heart failure", "pneumonia")){
    stop('invalid outcome')
    } else {
    si <- which(data2[, "state"] == state)
    ts <- data2[si, ]    # extracting data for the called state
    oi <- as.numeric(ts[, eval(outcome)])
    min_val <- min(oi, na.rm = TRUE)
    result  <- ts[, "hospital"][which(oi == min_val)]
    output  <- result[order(result)]
  }
  return(output)
}
best("TX", "heart attack")
best("TX", "heart failure")
best("MD", "heart attack")
best("MD", "pneumonia")
# Assignment 3
best("SC", "heart attack")
best("NY", "pneumonia")
best("AK", "pneumonia")
```

# Part 3: Ranking hospitals by outcome in a state

```{r}
# create a rankhospital function
rankhospital <- function(state, outcome, rank = "best"){
  ## Read outcome data
  data <- read.csv("hosp_compare/outcome-of-care-measures.csv", colClasses = "character")
  data2   <- as.data.frame(cbind(data[, 2],  # hospital
                              data[, 7],  # state
                              data[, 11],  # heart attack
                              data[, 17],  # heart failure
                              data[, 23]), # pneumonia
                        stringsAsFactors = FALSE)
  colnames(data2) <- c("hospital", "state", "heart attack", "heart failure", "pneumonia")
  
  ## Check that state and outcome are valid
  if (!state %in% data2[, "state"]) {
    stop('invalid state')
  } else if (!outcome %in% c("heart attack", "heart failure", "pneumonia")){
    stop('invalid outcome')
  } else if (is.numeric(rank)) {
    si <- which(data2[, "state"] == state)
    ts <- data2[si, ]                     # extracting dataframe for the called state
    ts[, eval(outcome)] <- as.numeric(ts[, eval(outcome)])
    ts <- ts[order(ts[, eval(outcome)], ts[, "hospital"]), ]
    output <- ts[, "hospital"][rank]
  } else if (!is.numeric(rank)){
    if (rank == "best") {
      output <- best(state, outcome)
    } else if (rank == "worst") {
      si <- which(data2[, "state"] == state)
      ts <- data2[si, ]    
      ts[, eval(outcome)] <- as.numeric(ts[, eval(outcome)])
      ts <- ts[order(ts[, eval(outcome)], ts[, "hospital"], decreasing = TRUE), ]
      output <- ts[, "hospital"][1]
    } else {
      stop('invalid rank')
    }
  }
  return(output)
}
# try out
rankhospital("TX", "heart failure", 4)
rankhospital("MD", "heart attack", "worst")
rankhospital("MN", "heart attack", 5000)
# Assignment 3
rankhospital("NC", "heart attack", "worst")
rankhospital("WA", "heart attack", 7)
rankhospital("TX", "pneumonia", 10)
rankhospital("NY", "heart attack", 7)
```


# Part 4: Ranking hospitals in all states

```{r}
rankall <- function(outcome, num = "best"){
  ## Read outcome data
  data <- read.csv("hosp_compare/outcome-of-care-measures.csv", colClasses = "character")
  data2   <- as.data.frame(cbind(data[, 2],  # hospital
                              data[, 7],  # state
                              data[, 11],  # heart attack
                              data[, 17],  # heart failure
                              data[, 23]), # pneumonia
                        stringsAsFactors = FALSE)
  colnames(data2) <- c("hospital", "state", "heart attack", "heart failure", "pneumonia")
  data2[, eval(outcome)] <- as.numeric(data2[, eval(outcome)])
  
  ## Check that state and outcome are valid
  
  if (!outcome %in% c("heart attack", "heart failure", "pneumonia")){
    stop('invalid outcome')
  } else if (is.numeric(num)) {
    by_state <- with(data2, split(data2, state))
    ordered  <- list()
    for (i in seq_along(by_state)){
      by_state[[i]] <- by_state[[i]][order(by_state[[i]][, eval(outcome)], 
                                           by_state[[i]][, "hospital"]), ]
      ordered[[i]]  <- c(by_state[[i]][num, "hospital"], by_state[[i]][, "state"][1])
    }
    result <- do.call(rbind, ordered)
    output <- as.data.frame(result, row.names = result[, 2], stringsAsFactors = FALSE)
    names(output) <- c("hospital", "state")
  } else if (!is.numeric(num)) {
    if (num == "best") {
      by_state <- with(data2, split(data2, state))
      ordered  <- list()
      for (i in seq_along(by_state)){
        by_state[[i]] <- by_state[[i]][order(by_state[[i]][, eval(outcome)],
                                             by_state[[i]][, "hospital"]), ]
        ordered[[i]]  <- c(by_state[[i]][1, c("hospital", "state")])
      }
      result <- do.call(rbind, ordered)
      output <- as.data.frame(result, stringsAsFactors = FALSE)
      rownames(output) <- output[, 2]
    } else if (num == "worst") {
      by_state <- with(data2, split(data2, state))
      ordered  <- list()
      for (i in seq_along(by_state)){
        by_state[[i]] <- by_state[[i]][order(by_state[[i]][, eval(outcome)],
                                             by_state[[i]][, "hospital"], 
                                             decreasing = TRUE), ]
        ordered[[i]]  <- c(by_state[[i]][1, c("hospital", "state")])
      }
      result <- do.call(rbind, ordered)
      output <- as.data.frame(result, stringsAsFactors = FALSE)
      rownames(output) <- output[, 2]
    } else {
      stop('invalid num')
    }
  }
  return(output)
}
# try out
head(rankall("heart attack", 20), 10)
tail(rankall("pneumonia", "worst"), 3)
tail(rankall("heart failure"), 10)
# Assignment 3
# 8
r <- rankall("heart attack", 4)
as.character(subset(r, state == "HI")$hospital)
# 9
r <- rankall("pneumonia", "worst")
as.character(subset(r, state == "NJ")$hospital)
# 10
r <- rankall("heart failure", 10)
as.character(subset(r, state == "NV")$hospital)
```









